---
title: "Exploring Storms"
author: "Pushpendu Biswas"
execute:
  echo: false
format:
  html:
    theme: cosmo
    code-fold: true
---

```{r}
#| label: setup
#| include: false
suppressPackageStartupMessages(library(tidyverse))
```

## Introduction

This analysis explores the historical data on storms from the `dplyr::storms` dataset, which includes tracking data for tropical storms and hurricanes.

## Named Storms Per Year

Below is a visualization showing the trend in the number of unique named storms recorded each year.

```{r}
#| label: plot-storms-per-year
#| fig-cap: "Annual number of unique named storms."
#| warning: false
#| message: false

# Calculate number of named storms per year by status
storms_per_year <- storms %>%
  filter(status %in% c("tropical depression", "tropical storm", "hurricane")) %>%
  mutate(status = factor(status, levels = c("tropical depression", "tropical storm", "hurricane"))) %>%
  distinct(year, name, status) %>%
  count(year, status, name = "count") %>%
  complete(year, status, fill = list(count = 0))

# Plot using ggplot2
ggplot(storms_per_year, aes(x = year, y = count, fill = status)) +
  geom_area(position = "stack", alpha = 0.85, color = "white", linewidth = 0.1) +
  scale_fill_manual(
    values = c(
      "tropical depression" = "#8cc5e3",
      "tropical storm"      = "#f4a261",
      "hurricane"           = "#b03a2e"
    ),
    labels = c(
      "tropical depression" = "Tropical Depression",
      "tropical storm"      = "Tropical Storm",
      "hurricane"           = "Hurricane"
    )
  ) +
  labs(
    title = "Storm Frequency by Type (1975 - 2022)",
    subtitle = "Annual counts of unique tropical depressions, tropical storms, and hurricanes",
    x = "Year",
    y = "Number of Unique Storms",
    fill = "Storm Status",
    caption = "Source: dplyr::storms"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold", size = 16, color = "#2c3e50"),
    plot.subtitle = element_text(size = 12, color = "#7f8c8d"),
    legend.position = "bottom",
    panel.grid.minor = element_blank(),
    plot.caption = element_text(face = "italic", size = 9, color = "#bdc3c7")
  )
```

## Summary Statistics

Over the period from `r min(storms$year)` to `r max(storms$year)`, there were a total of `r storms %>% distinct(year, name) %>% nrow()` unique storms tracked. The year with the highest number of named storms was `r storms %>% distinct(year, name) %>% count(year) %>% filter(n == max(n)) %>% pull(year)` with `r storms %>% distinct(year, name) %>% count(year) %>% filter(n == max(n)) %>% pull(n)` storms.
