---
title: "Exploring Storms"
author: "Pushpendu Biswas"
execute:
  echo: false
format:
  html:
    theme: cosmo
    code-fold: true
---

```{r}
#| label: setup
#| include: false
suppressPackageStartupMessages(library(tidyverse))
```

## Introduction

This analysis explores the historical data on storms from the `dplyr::storms` dataset, which includes tracking data for tropical storms and hurricanes.

## Named Storms Per Year

Below is a visualization showing the trend in the number of unique named storms recorded each year.

```{r}
#| label: plot-storms-per-year
#| fig-cap: "Annual number of unique named storms."
#| warning: false
#| message: false

# Calculate number of named storms per year
storms_per_year <- storms %>%
  distinct(year, name) %>%
  count(year, name = "num_storms")

# Plot using ggplot2
ggplot(storms_per_year, aes(x = year, y = num_storms)) +
  geom_line(color = "#1f77b4", linewidth = 1) +
  geom_point(color = "#0f3d5f", size = 2) +
  geom_smooth(method = "loess", color = "#ff7f0e", fill = "#ff7f0e", alpha = 0.1, formula = y ~ x) +
  labs(
    title = "Named Storms Per Year (1975 - 2022)",
    subtitle = "Tracking the frequency of tropical storms and hurricanes over time",
    x = "Year",
    y = "Number of Unique Storms",
    caption = "Source: dplyr::storms"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold", size = 16, color = "#2c3e50"),
    plot.subtitle = element_text(size = 12, color = "#7f8c8d"),
    panel.grid.minor = element_blank(),
    plot.caption = element_text(face = "italic", size = 9, color = "#bdc3c7")
  )
```

## Summary Statistics

Over the period from `r min(storms$year)` to `r max(storms$year)`, there were a total of `r storms %>% distinct(year, name) %>% nrow()` unique storms tracked. The year with the highest number of named storms was `r storms %>% distinct(year, name) %>% count(year) %>% filter(n == max(n)) %>% pull(year)` with `r storms %>% distinct(year, name) %>% count(year) %>% filter(n == max(n)) %>% pull(n)` storms.
